---
title: "Staircase 1PLUS Model"
author: "Edgar Santos-Fernández"
date: "3/12/2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using a 1PLUS model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document outlines the fitting of a one-parameter spatial item response
 model (1PLUS), which involves abilities and difficulties associated with the
 species and spatial locations. However, it does not includes guessing and
 slope parameters. See, more details on the ir_spat function help. We use a
 exponential covariance matrix (spat_model = ‘exp’).


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Load Requisite Libraries
``` {r loadlibraries, results='hide'}
library('Rcpp')
library('dplyr')
library('ggplot2')
library('rstan')
library('bayesplot')
library('ggrepel')
library('RColorBrewer')
library('rgeos')
library('dismo')
library('ggvoronoi')
library('staircase')
```

### Prepare data and set options
NOTE: using options(mc.cores = parallel::detectCores()) is optimal.
CRAN package vignettes limit the maximum number of cores to 2.
```{r,eval=TRUE}
diverdata <- staircase::reefdiverdata

rstan_options(auto_write = TRUE)
options(mc.cores = 2) #options(mc.cores = parallel::detectCores()) is

seed <- 202105
set.seed(seed)
```

### Trim dataset
``` {r, eval=TRUE, results='hide'}
diverdata <- diverdata[diverdata$siteID %in% names(table(diverdata$siteID)[(table(diverdata$siteID) > 100) ]),]
diverdata <- diverdata[diverdata$userID %in% names(table(diverdata$userID)[(table(diverdata$userID) > 30) ]),]
diverdata <- diverdata[diverdata$userID %in% unique(diverdata$userID)[1:30],]

images <- sort(table(diverdata$siteID), decreasing = T)
images <- names(images[images < 31])
diverdata <- diverdata[diverdata$siteID %in% images,]
``` 

### Create unique user IDs and site IDs
``` {r, eval=TRUE}
diverdata$siteID <- as.numeric(factor(diverdata$siteID))
diverdata$userID <- as.numeric(factor(diverdata$userID))
```

### Examine the observation rate of the users in the smaller dataset
This smaller dataset features 43 images and 13 users' classifications.
```{r}
ggplot(diverdata) + 
  geom_raster(aes(y= siteID, x = userID)) + 
  ylab('Site ID') +
  xlab('User ID') +
  theme_bw()
```

### Run model
Then using the function ir_spat, we fit a one-parameter spatial item response 
model (1PLUS), which involves abilities and difficulties associated with the 
species and spatial locations. However, it does not includes guessing and 
slope parameters.

The number of iterations used in this example is small. A more
accurate output could be obtained by increasing the number of iterations.
However, this will significantly increase the run time of this code.

This prepared sample will take about 4 minutes to run to completion.

Note: If the following code is run on a computer with a single core, or a computer than
cannot use parallel operations the code will take significantly longer to run.
``` {r, eval=TRUE, results='hide'}
fit <- ir_spat(formula = siteID ~ -1 + hard_bin, # covariates affecting the difficulty
               data = diverdata, # a data frame
               spat_model = 'exp', # spatial covariance matrix
               itemtype = '1PLUS', # item response model
               abil = 'userID', # participants ids
               diff = 'siteID', # location id
               y = 'correct', # binary response variable
               coords = c("lon", "lat"), # coordinates
               iter = 400,
               warmup = 200, 
               chains = 3,
               refresh = 5,
               seed = seed
)
```


### Computing Summary Statistics
```{r, eval=TRUE}
stats <- summary(fit)
stats <- stats$summary
array <- as.array(fit)
```


### Comparing and evaluating participants abilities
``` {r}
abils <- stats[grep('abil\\[', rownames(stats)),]
abils <- cbind(abils, diverdata %>% group_by(userID) %>% summarize(ns = n(), prop = mean(correct)) %>% data.frame())
names(abils)[grep('%', names(abils))] <- c('q2.5','q25', 'q50', 'q75', 'q97.5')

ggplot(abils, aes(x= prop, y = mean)) + geom_point() +
  geom_text_repel(aes(x= prop, y = mean, label = userID), size = 2.5) +
  xlab('Proportion of correct classification')+
  ylab('Abilities')
```

### Plotting the relative difficulty of each site
``` {r}
diff <- stats[grep('difficulty\\[', rownames(stats)),]
diff <- cbind(diff, diverdata %>% group_by(siteID, lon, lat) %>% summarize(ns = n(), prop = mean(correct)) %>% data.frame())
diff$diff_cat <- cut(diff$mean, breaks = c(-10,quantile(diff$mean)[2:4],10), 1:4, include.lowest = T)
cols = brewer.pal(5,'Reds')
ggplot(diff , aes(lon, lat, fill = diff_cat)) +
  xlab("Longitude") +
  ylab("Latitude") +
  stat_voronoi(color="black") + scale_fill_manual(values=cols) +
  geom_point() +
  geom_text(aes(lon, lat, label = siteID), size = 2)+
  coord_fixed(ratio=1)+
  labs(fill = "Site difficulty")+
  theme_bw()
```

### Plotting the estimated species difficulties
``` {r}
species <- stats[grep('species\\[', rownames(stats)),]
species <- cbind(species, diverdata %>% group_by(True_Species) %>% summarize(ns = n(), prop = mean(correct)) %>% data.frame())
names(species)[grep('%', names(species))] <- c('q2.5','q25', 'q50', 'q75', 'q97.5')
ggplot(species, aes(x= prop, y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin=q2.5, ymax=q97.5))+
  xlab('Proportion of correct classification')+
  ylab('Species difficulty')
```

### Posterior distributions of the regression coefficients
``` {r}
mcmc_dens_overlay(
  array,
  pars = c(
    "beta[1]"),
  facet_args = list(nrow = 1)
)

```


